Stochastic Approximation Monte Carlo with a Dynamic Update Factor

### 2019-08-07 papers/histogram/sad-referee-response.md

...

Response to first referee

The first referee acknowledges that our paper is well written and that the field
is an important one, but does not believe that our work is a *'substantial
contribution*'' to the field.  The referee has three major criticisms of the
value of our work (aside from issues with the introduction section).

Our method is based on SAMC (as we clearly state in the paper) and is a substantial
improvement on this method (convergence is of 1 - 2 orders of magnitude superior
as can be seen from Fig. 3 and Fig. 4).  In addition, we also provide a detailed
comparison with 4 different methods on the square well system.

Furthermore, the guidelines that the referee is using presumably come from
Computational Physics Papers in the Physical Review E (2013). This specifically
states '...Rather, the requirement that "papers must contain new results in
physics" should be interpreted as a requirement of novelty with respect to the
method(s), the physics, or both.' As stated above it is only necessary that the
method is novel OR the physical problem is novel. As example, the paper below
accepted by PRE confirms this thought.

@@

**Multidimensional stochastic approximation Monte Carlo**
*(accepted by PRE)*
"the Wang-Landau MC [3] scheme and its mathematical generalization, stochastic
approximation MC (SAMC) [4–6].""

**Convergence estimation of flat-histogram algorithms based on simulation results**
*(accepted by Computer Physics Communications)*
"At the same time, Liang suggested a generalized simulation method:stochastic
approximation Monte Carlo (SAMC) [7,8] including the 1/t-WL as a particular case"

**Stochastic approximation Monte Carlo and Wang–Landau Monte Carlo applied to a
continuum polymer model**
*(accepted by Computer Physics Communications)*
"Liang et al.[30] showed that WLMC could be seen as a version of SAMC and using
the mathematical background of stochastic approximation methods they proved the
convergence of SAMC."

The referee further disputes the use of language *'we have introduced a new
algorithm'*. This language is commonly used to describe MC methods that have
their foundation in other methods (as SAD is a derivative of SAMC). However,
we did revise the paper to clarify in the conclusion that our algorithm is a
variant of the SAMC method.

@@

**----------------------------------------------------------------------------**

Minor details:

6. The referee asks why we chose a maximum temperature of infinity.  *We have
added a paragraph to Section III explaining the reasoning*, which is that
infinite temperature allows the system to randomize more reliably, and costs us
essentially nothing.  We do note that for some systems or situations in which
the energy of maximum entropy is very high (or worse, infinite) we would need
to introduce a maximum temperature.

7. On page 5 the authors say "SAD should perform similarly ...". Is
there any evidence?
**We have edited the paper to make clear that we have not tested and confirmed
the behavior of SAD in systems with a continuum of energies.**

@@

**----------------------------------------------------------------------------**

1. The referee wrote that the description of the flat histogram methods in
section II was confusing, and should be rewritten to highlight the differences
between methods.  We have added ...
**We have better described the implementation of each method in more detail in
section II to highlight the major differences in implementation between the
methods. The update factor is now shown for each method so that it can be seen
how each method differs.**

**----------------------------------------------------------------------------**

2. The referee shared that the entropy argument in Section II.D is unclear, and
wished for the relationship between entropy and the update factor to be more
fully described, as well as the basis for Equation 7.
**We have added an additional description of the relationship between entropy,
ln w, and the update factor immediately after equation 7 to make more clear how
the total change in entropy connects with the update factor schedule.**

@@

**----------------------------------------------------------------------------**

3. The referee asked for a better choice of line styles to highlight the SAD
method in our plots.
**We found that by thickening the SAD line style and thinning SAMC lines we
could make it considerably easier to distinguish between the different methods
and to better highlight the SAD method in our plots.**

### 2019-08-07 papers/histogram/sad.tex
changes to paper in response to referees

\begin{abstract}
  We present a new Monte Carlo algorithm based on the Stochastic
  Approximation Monte Carlo (SAMC) algorithm for directly calculating
  the density of states. The proposed method is Stochastic
  Approximation with a Dynamic update factor (SAD)
  which dynamically adjusts the update factor $\gamma_t$ during the course of
  the simulation. We test this method on the square-well fluid and
  compare the convergence time and average entropy error for several
  related
  Monte Carlo methods. We find that both the SAD and $1/t$-Wang-Landau ($1/t$-WL)
  methods rapidly converge to the
  correct density of states without the need for the user to specify an
  arbitrary tunable parameter $t_0$ as in the case of SAMC.  SAD requires
  as input the temperature range of interst, in contrast to
  $1/t$-WL, which requires that the user identify the interesting range
  of energies.  Thus SAD is more convenient when the range of energies is
  not known in advance.
\end{abstract}

@@

\section{Introduction}
Over the past several decades a number of flat histogram Monte Carlo simulation
algorithms have been developed which calculate the thermodynamic properties of
various systems for over a range of temperatures.  This development began with
the original histogram method, which used a single canonical Monte Carlo
simulation to predict properties for nearby
temperatures~\cite{ferrenberg1988new}.  For large systems this approach is
limited to a narrow temperature range because a single canonical simulation
explores only a small range of energies. Berg and Neuhaus developed multicanonical
methods that introduced a weight function to enable flat histogram sampling
which improved the exploration of configuration space and allows the system
to
overcome free energy barriers~\cite{berg1991multicanonical, berg1992multicanonical}.
These works led to an explosion in the interest of
``flat'' (or
``broad'') histogram methods~\cite{penna1996broad, penna1998broad,
swendsen1999transition, wang2001determining, wang2001efficient,
trebst2004optimizing}, which explore a wider range of energies.  Another
benefit, in contrast with low-temperature canonical Monte Carlo, is that these
approaches cannot be trapped in a local energy minimum.

Wang and Landau introduced the most widely used flat histogram algorithm (WL)
that uses an update factor and a statistical histogram to compute the density of
states of a given system~\cite{wang2001determining, wang2001efficient}.  While
the method is incredibly powerful, it has a few disadvantages. Firstly, it
requires the user to select the range of energies to be
studied~\cite{landau2004new}. This adds an additional hurdle to its application
to systems for which the interesting range of energies is not known \emph{a
priori}.  The simulation violates detailed balance, albeit briefly as the size of
the violation decreases with time, which complicates convergence analysis.  In
fact, the error in a WL computation has been demonstrated to saturate
at a non-zero value~\cite{yan2003fast}, i.e. the method does not converge to the
true density of states~\cite{belardinelli2007wang, belardinelli2007fast,
belardinelli2008analysis, belardinelli2014intrinsic, singh2012density}.
%

@@

Liang began to consider that WL could be considered a special case of the
Stochastic Approximation approach, whose convergence could be mathematically
proven~\cite{liang2006theory}. In 2007, Liang, Liu, and Carrol \emph{et al.}
developed Stochastic Approximation Monte Carlo
(SAMC)~\cite{liang2007stochastic}, and proved its convergence, although the
method still has a system specific user-defined parameter which must be tuned
when applying this algorithm to a new system.

Kim \emph{et al.} introduced Statistical Temperature Monte Carlo (STMC) and
the related Statistical Temperature Molecular Dynamics (STMD) which is an
improvement on the WL method which approximates the entropy (or log density of states)
as a piecewise
linear function, which improves convergence for systems with a continuously varying
energy~\cite{kim2006statistical, kim2007statistical}. STMC
applied to WL requires a temperature range be specified rather than an energy
range.  Kim \emph{et al.} extended this work as Replica Exchange Statistical
Temperature Monte Carlo (RESTMC), which uses replica exchange of multiple overlapping
STMC simulations to improve convergence~\cite{kim2009replica}.
More recently, Junghans \emph{et al.} have demonstrated a close
connection between metadynamics, which was introduced by Laio and
Parinello~\cite{laio2002escaping}, and WL-based Monte Carlo
methods, with STMD forging the connection~\cite{junghans2014molecular}.

@@

\section{Flat histogram methods}\label{sec:histogram}
\jpsays{REFEREE 2: The description of flat histogram methods in section II was a bit hard
to understand at first and may need to be edited to better highlight
the differences between the methods. Also, the entropy argument in
section II.D is unclear. The relationship between entropy, the
histogram, and the update factor needs to be more fully described and
the basis for equation 7 (and the approximations made there) needs to
be made clearer.}

The goal of flat histogram methods (also called \emph{broad histogram}
or \emph{multicanonical} methods) is to simulate each energy with
similar accuracy, so as to accurately determine the density of states
over a broad range of energies---and thus to determine the thermodynamic
quantities such as heat capacity or internal energy
over a broad range of temperatures.
Properties that require more information---such as a spatial
correlation function or a response function---can still be computed
for any temperature, provided statistics are collected for each
individual energy, which can then be reweighted for any
temperature~\cite{panagiotopoulos1998phase, panagiotopoulos2000monte,
errington2003direct}.

@@

The Wang-Landau approach~\cite{wang2001efficient,wang2001determining,
  landau2014guide} begins with $\gamma_t=1$, and then decreases $\gamma$
in discrete stages.  When the energy histogram is sufficiently flat,
$\gamma_t$ is decreased by a specified factor of $\frac12$.  The
flatness is defined by the ratio between the minimum value of the
histogram and its average value.  When this flatness reaches a
specified threshold (typically 0.8), the $\gamma_t$ value is decreased
and the histogram is reset to zero.  This approach requires that the
energy range of interest be known in advance, and difficulties can
occur with this flatness criteria due to the fact that some energies
in this energy range might never be
sampled~\cite{troster2005wang}.  The entire process is repeated
until $\gamma_t$ reaches a desired cutoff.

The Wang-Landau approach thus has three parameters that need be
specified: the factor by which to decrease $\gamma_t$ when flatness is
acheived, the flatness criterion, and the cutoff that determines when
the computation is complete.  In addition, an energy range (or in
general, a \emph{set} of energies) must be supplied, so that the
flatness criterion can be defined.

\jpsays{remove and add to intro as this does not convey how to implement WL!}
While this approach is very efficient and has been widely used, it
suffers a few shortcomings.  Firstly, the set of energies must be
specified ~\cite{wang2001efficient, schulz2003avoiding, yan2003fast},
which may require multiple simulations. Secondly, while Wang-Landau
converges quickly, it does not in general converge to the true density
of states ~\cite{belardinelli2008analysis, zhou2008optimal}.  It can
and does decrease $\gamma_t$ so quickly that it will never (for any
cutoff value) decrease the error in the density of states beyond a
given nonzero value.

@@

The $1/t$-WL algorithm ensures convergence by preventing the $\gamma_t$ factor
from dropping below $N_S/t$~\cite{belardinelli2008analysis, schneider2017convergence}.
The method follows the standard WL algorithm with two modifications.  Firstly, the histogram
is considered flat, and $\gamma_t$ is decreased by a factor of two, when every
energy state has been visited once, i.e. when the WL ``flatness'' becomes nonzero.
Secondly, when the $\gamma_t$ becomes less than $N_S/t$, the algorithm switches
to use
\begin{align}
  \gamma_t = \begin{cases}
     \gamma_{WL} & \gamma_{WL} > \frac{N_S}{t} \\
     \frac{N_S}{t} & \text{forever more}
 \end{cases}
\end{align}
where $t$ is the number of moves and $N_S$ is the number of energy bins.
\jpsays{David and I got to here!}

\subsection{SAMC}
The Stochastic Approximation Monte Carlo (SAMC) algorithm addresses
the lack of convergence of Wang-Landau's approach with a simple
schedule by which the update factor $\gamma_t$ is continuously
decreased~\cite{liang2007stochastic, werlich2015stochastic,
  schneider2017convergence}.  The update factor is defined in the
original implementation~\cite{liang2007stochastic} in terms of an
arbitrary tunable parameter $t_0$,
\begin{align}
\gamma_{t}^{\text{SA}} =\frac{t_0}{\max(t_0,t)}\label{eq:1}
\end{align}
where as above $t$ is the number of moves that have been attempted.
SAMC offers extreme simplicity, combined with is proven convergence.
Provided the update factor satisfies
\begin{align}
\sum_{t=1}^\infty \gamma_{t} = \infty \quad\textrm{and}\quad
\sum_{t=1}^\infty \gamma_{t}^\zeta < \infty
\end{align}
where $\zeta \in \{1,2\}$, Liang has shown that the weights are proven
to converge to the true density of states~\cite{liang2006theory,
  liang2007stochastic}.  In addition, the energy range need not be
known a priori.  The time to converge depends only on the choice of
parameter $t_0$.  Unfortunately, there is no prescription for finding
an acceptable value for $t_0$, and while the algorithm formally
converges, for a poor choice of $t_0$ that convergence can be far too
slow to be practical. Liang \emph{et al.} give a rule of thumb in
which $t_0$ is chosen in the range from $2N_S$ to $100N_S$ where $N_S$
is the number of energy bins~\cite{liang2007stochastic}.  Schneider
\emph{et al.} found that for the Ising model this heuristic is helpful
for small spin systems, but that larger systems require an even higher
$t_0$ value~\cite{schneider2017convergence}.  We will describe below
one case we examined, in which $t_0$ needs to be as much as two orders
of magnitude higher than the rule of thumb of $100N_S$ in order to
converge in $10^{12}$ moves.

Werlich \emph{et al.} proposed scaling the SAMC $\gamma_t$ by a factor
$\gamma_0$~\cite{werlich2015stochastic}.  While this may result in an
improved rate of convergence, it adds yet another parameter that must
be empirically determined, and we have not explored this degree of
freedom.

@@

We can place a rigorous \emph{lower} bound $t_{\min}$ on the number of
moves required to find the true entropy by considering the total change
that needs to be made to the entropy.
\begin{align}
  \Delta S_{\text{tot}} \equiv \sum_E S(E) - S_{\min}
\end{align}
The minimum number of moves that could converge the entropy is
the number of moves that will enable the above total entropy change,
which we can approximate using an integral:
\begin{align}
   \Delta S_{\text{tot}} &= \sum_{t=0}^{t_{\min}} \gamma_t^{SA} \\
  % &= \int_0^{t_{\min}} \gamma(t)dt \\
  &= t_0 + \int_{t_0}^{t_{\min}} \frac{t_0}{t}dt
  \\
  &= t_0\left(1 + \ln\left(\frac{t_{\min}}{t_0}\right)\right)
\end{align}
Solving for ${t_{\min}}$ we find that
\begin{align}
  {t_{\min}} &= t_0 e^{\frac{\Delta S_{\text{tot}}}{t_0} - 1}
\end{align}
which means that the minimum time to convergence grows exponentially
as $t_0$ is made smaller.  You \emph{seriously} don't want to underestimate
$t_0$!

One might reasonably choose to err by selecting a large $t_0$.
The rate of convergence is harder to estimate when $t_0$ is large, but
in general $\gamma_t$ itself forms a lower bound on the accuracy with which
the entropy may be known, with an unknown prefactor which is roughly
the coherence time of the Monte Carlo simulation \jpsays{Michael Perlin asked about this in his comments and my understanding was that we meant the time for $\gamma_t$ to reach 1/t behavior. Since this is part of a section needing editing I thought we should clarify}.  Since $\gamma_t$ is
given by $t_0/t$, the time to converge to a given accuracy is increased
in proportion to the ratio by which we overestimate $t_0$.  Thus, while
it is exponentially painful to underestimate $t_0$, overestimating by
several orders of magnitude is also not acceptable.  We should note
that these extreme limiting cases do not preclude the possibility that there
is a wide range of $t_0$ values that lead to an acceptable convergence
rate.

@@

\begin{figure}
  \includegraphics[width=\columnwidth]{gamma-n256}
  \caption{The update factor $\gamma_t$ versus iteration for $N=256$
    square-well atoms, for four different methods: WL, $1/t$-WL, SAMC,
    and SAD}
    \label{fig:gamma-vs-t}
\end{figure}
Figure~\ref{fig:gamma-vs-t} compares $\gamma_t$ for the related
methods SAD, WL, $1/t$-WL, and SAMC.  For SAMC, $\gamma_t$ remains
constant before dropping as $1/t$.  WL $\gamma_t$ remains at 1 for
many iterations, and then decreases very rapidly, with $1/t$-WL
behaving similarly but decreasing more aggressively before
transitioning to a more conservative $1/t$ behavior.  The update
factor for SAD fluctuates dynamically around a value less than 1 for
early MC moves, and then decreases as approximately $1/t$ while
continuing to fluctuate as new energies are found to be important.  At
intermediate times, the SAD $\gamma_t$ decreases as $1/t^2$ before
asymptoting to $N_S/t$, which is the same as $1/t$-WL.

@@

A significant advantage of SAD over SAMC---which the $1/t$-WL and WL
methods share \emph{after} they have found all the energies---is that
the schedule for $\gamma_t$ automatically responds to the choice of bin
size. SAD should perform similarly over a wide range of bin sizes
because $\gamma_t \propto \nicefrac{N_S}{t}$.  As the number of energy
states $N_S$ found increases (fine binning), the time spent $t$ in
each bin will decrease with the effect that the convergence should be
roughly independent of the bin size chosen.  SAMC could be used with a
prefactor $\gamma_0$ to aid in a similar
way~\cite{werlich2015stochastic} but this adds yet another parameter
for the user to choose.

@@

\section{Conclusions}
\jpsays{We have introduced SAD, a new variant of SAMC, that}
We have introduced a new algorithm that effectively samples the
energy space corresponding to a desired range of temperatures for a
variety of system sizes.
%
We find that both SAD and $1/t$-WL demonstrate excellent and robust
convergence.
They both converge more rapidly than SAMC, and unlike WL consistently
converge to the correct density of states. SAD requires
the user to specify a temperature range of interest rather than an
energy range of interest as $1/t$-WL does.  For use cases in which
a range of desired temperatures is known, this will make the SAD method
considerably more convenient.

>>>>

\begin{abstract}
  We present a new Monte Carlo algorithm based on the Stochastic
  Approximation Monte Carlo (SAMC) algorithm for directly calculating
  the density of states. The proposed method is Stochastic
  Approximation with a Dynamic update factor (SAD)
  which dynamically adjusts the update factor $\gamma_t$ during the course of
  the simulation. We test this method on the square-well fluid and
  compare the convergence time and average entropy error for several
  related
  Monte Carlo methods. We find that both the SAD and $1/t$-Wang-Landau ($1/t$-WL)
  methods rapidly converge to the
  correct density of states without the need for the user to specify an
  arbitrary tunable parameter $t_0$ as in the case of SAMC.  SAD requires
  as input the temperature range of interst, in contrast to
  $1/t$-WL, which requires that the user identify the interesting range
  of energies.  Thus SAD is more powerful in the common case in which the range
  of energies is not known in advance.
\end{abstract}

@@

\section{Introduction}
Over the past several decades a number of flat histogram Monte Carlo simulation
algorithms have been developed which calculate the thermodynamic properties of
various systems for over a range of temperatures.  This development began with
the original histogram method, which used a single canonical Monte Carlo
simulation to predict properties for nearby
temperatures~\cite{ferrenberg1988new}.  For large systems this approach is
limited to a narrow temperature range because a single canonical simulation
explores only a small range of energies. Berg and Neuhaus developed multicanonical
methods that introduced a weight function to enable flat histogram sampling
which improved the exploration of configuration space and allows the system
to
overcome free energy barriers~\cite{berg1991multicanonical, berg1992multicanonical}.
These works led to an explosion in the interest of
``flat'' (or
``broad'') histogram methods~\cite{penna1996broad, penna1998broad,
swendsen1999transition, wang2001determining, wang2001efficient, landau2004new, schulz2003avoiding, yan2003fast,
trebst2004optimizing,
belardinelli2007wang, belardinelli2007fast,
belardinelli2008analysis, belardinelli2014intrinsic, singh2012density, zhou2008optimal,
schneider2017convergence, liang2006theory, liang2007stochastic,
kim2006statistical,
kim2007statistical, kim2009replica, junghans2014molecular}, which explore a wider range of energies.  Another
benefit, in contrast with low-temperature canonical Monte Carlo, is that these
approaches cannot be trapped in a local energy minimum.

Wang and Landau introduced the most widely used flat histogram algorithm (WL)
that uses an update factor and a statistical histogram to compute the density of
states of a given system~\cite{wang2001determining, wang2001efficient}.  While
the method is incredibly powerful, it has a few disadvantages. Firstly, it
requires the user to select the range of energies to be
studied~\cite{landau2004new, wang2001efficient, schulz2003avoiding, yan2003fast}. This adds an additional hurdle to its application
to systems for which the interesting range of energies is not known \emph{a
priori}.  The simulation violates detailed balance, albeit briefly as the size of
the violation decreases with time, which complicates convergence analysis.  In
fact, the error in a WL computation has been demonstrated to saturate
at a non-zero value~\cite{yan2003fast}, i.e. the method does not converge to the
true density of states~\cite{belardinelli2007wang, belardinelli2007fast,
belardinelli2008analysis, belardinelli2014intrinsic, singh2012density, zhou2008optimal}.
%

@@

Liang began to consider that WL could be considered a special case of the
Stochastic Approximation approach, whose convergence could be mathematically
proven~\cite{liang2006theory}. In 2007, Liang, Liu, and Carrol \emph{et al.}
developed Stochastic Approximation Monte Carlo
(SAMC)~\cite{liang2007stochastic}, and proved its convergence, although the
method still has a system specific user-defined parameter which must be tuned
when applying this algorithm to a new system.  In
contrast to the WL-based methods, the SAMC method does not require users to
determine the energy range of interest \emph{a priori}.

Kim \emph{et al.} introduced Statistical Temperature Monte Carlo (STMC) and the
related Statistical Temperature Molecular Dynamics (STMD) which is an
adaption of the WL method which approximates the entropy (or log density of
states) as a piecewise linear function, which improves convergence for systems
with a continuously varying energy~\cite{kim2006statistical,
kim2007statistical}. STMC applied to WL requires a temperature range be
specified rather than an energy range.  Kim \emph{et al.} extended this work as
Replica Exchange Statistical Temperature Monte Carlo (RESTMC), which uses
replica exchange of multiple overlapping STMC simulations to improve
convergence~\cite{kim2009replica}. More recently, Junghans \emph{et al.} have
demonstrated a close connection between metadynamics, which was introduced by
Laio and Parinello~\cite{laio2002escaping}, and WL-based Monte Carlo methods,
with STMD forging the connection~\cite{junghans2014molecular}.

@@

\section{Flat histogram methods}\label{sec:histogram}
The goal of flat histogram methods (also called \emph{broad histogram}
or \emph{multicanonical} methods) is to simulate each energy with
similar accuracy, so as to accurately determine the density of states
over a broad range of energies---and thus to determine the thermodynamic
quantities such as heat capacity or internal energy
over a broad range of temperatures.
Properties that require more information---such as a spatial
correlation function or a response function---can still be computed
for any temperature, provided statistics are collected for each
individual energy, which can then be reweighted for any
temperature~\cite{panagiotopoulos1998phase, panagiotopoulos2000monte,
errington2003direct}.

@@

The Wang-Landau approach~\cite{wang2001efficient,wang2001determining,
  landau2014guide} begins with $\gamma^{\text{WL}}_{t=0}=1$, and then decreases $\gamma^{\text{WL}}$
in discrete stages.  When the energy histogram is sufficiently flat,
$\gamma^{\text{WL}}$ is decreased by a specified factor of $\frac12$.  The
flatness is defined by the ratio between the minimum value of the
histogram and its average value.  When this flatness reaches a
specified threshold (typically 0.8), the $\gamma^{\text{WL}}$ value is decreased
and the histogram is reset to zero.  This approach requires that the
energy range of interest be known in advance, and difficulties can
occur with this flatness criteria due to the fact that some energies
in this energy range might never be
sampled~\cite{troster2005wang}.  The entire process is repeated
until $\gamma^{\text{WL}}$ reaches a desired cutoff.

The Wang-Landau approach thus has three parameters that need be
specified: the factor by which to decrease $\gamma^{\text{WL}}$ when flatness is
acheived, the flatness criterion, and the cutoff that determines when
the computation is complete.  In addition, an energy range (or in
general, a \emph{set} of energies) must be supplied, so that the
flatness criterion can be defined.

@@

The $1/t$-WL algorithm ensures convergence by preventing the $\gamma_t$ factor
from dropping below $N_S/t$~\cite{belardinelli2008analysis, schneider2017convergence}.
The method follows the standard WL algorithm with two modifications.  Firstly, the histogram
is considered flat, and $\gamma_t$ is decreased by a factor of two, when every
energy state has been visited once, i.e. when the WL ``flatness'' becomes nonzero.
Secondly, when $\gamma^{\text{WL}} < N_S/t$ at time $t_0$, the algorithm switches
to use  $\gamma_t = N_S/t$ for the remainder of the simulation:
\begin{align}
  \gamma_t^{1/t\text{-WL}} = \begin{cases}
     \gamma^{\text{WL}}_t & \gamma^{\text{WL}}_t > \frac{N_S}{t} \\
     \frac{N_S}{t} & t \ge t_0
 \end{cases}
\end{align}
where $t$ is the number of moves, $\gamma^{\text{WL}}_t$ is the Wang-Landau update factor
at move $t$, and $N_S$ is the number of energy bins.

\subsection{SAMC}
The Stochastic Approximation Monte Carlo (SAMC) algorithm addresses
the lack of convergence of Wang-Landau's approach with a simple
schedule by which the update factor $\gamma_t$ is continuously
decreased~\cite{liang2007stochastic, werlich2015stochastic,
  schneider2017convergence}.  The update factor is defined in the
original implementation~\cite{liang2007stochastic} in terms of an
tunable parameter $t_0$,
\begin{align}
\gamma_{t}^{\text{SA}} =\frac{t_0}{\max(t_0,t)}\label{eq:1}
\end{align}
where as above $t$ is the number of moves that have been attempted.
SAMC offers extreme simplicity, combined with is proven convergence.
Provided the update factor satisfies
\begin{align}
\sum_{t=1}^\infty \gamma_{t} = \infty \quad\textrm{and}\quad
\sum_{t=1}^\infty \gamma_{t}^\zeta < \infty
\end{align}
where $\zeta \in \{1,2\}$, Liang has proven that the weights converge to the
true density of states~\cite{liang2006theory,
  liang2007stochastic}.  In addition, the energy range need not be
known a priori.  The time to converge depends only on the choice of
parameter $t_0$.  Unfortunately, there is no prescription for finding
an acceptable value for $t_0$, and while the algorithm formally
converges, for a poor choice of $t_0$ that convergence can be far too
slow to be practical. Liang \emph{et al.} give a rule of thumb in
which $t_0$ is chosen in the range from $2N_S$ to $100N_S$ where $N_S$
is the number of energy bins~\cite{liang2007stochastic}.  Schneider
\emph{et al.} found that for the Ising model this heuristic is helpful
for small spin systems, but that larger systems require an even higher
$t_0$ value~\cite{schneider2017convergence}.  We will describe below
one case we examined, in which $t_0$ needs to be as much as two orders
of magnitude higher than the rule of thumb of $100N_S$ in order to
converge in $10^{12}$ moves.

Werlich \emph{et al.} proposed scaling the SAMC $\gamma_t$ by a factor
$\gamma_0$~\cite{werlich2015stochastic}.  While this may result in an
improved rate of convergence, it adds yet another parameter that must
be empirically determined, and we have not explored this additional degree of
freedom.

@@

We can place a rigorous \emph{lower} bound $t_{\min}$ on the number of
moves required to find the true entropy by considering the total change
that needs to be made to the entropy.
\begin{align}
  \Delta S_{\text{tot}} \equiv \sum_E S(E) - S_{\min}
\end{align}
This connects with $\gamma_t$ because the $\ln w(E)$ (our approximation for entropy)
has $\gamma_t$ added to it on each move.  Thus a minimum number of moves
that could possibly result in the true entropy $S(E)$ starting with a flat set of
weights is determined by the total entropy change required.  We can estimate
this number of moves required by summing the $\gamma_t$,
which we can approximate using an integral:
\begin{align}
   \Delta S_{\text{tot}} &= \sum_{t=0}^{t_{\min}} \gamma_t^{SA} \\
  % &= \int_0^{t_{\min}} \gamma(t)dt \\
  &\approx t_0 + \int_{t_0}^{t_{\min}} \frac{t_0}{t}dt
  \\
  &= t_0\left(1 + \ln\left(\frac{t_{\min}}{t_0}\right)\right)
\end{align}
Solving for ${t_{\min}}$ we find that
\begin{align}
  {t_{\min}} &= t_0 e^{\frac{\Delta S_{\text{tot}}}{t_0} - 1}
\end{align}
which means that the minimum time to converge grows exponentially
as $t_0$ is made smaller.  You \emph{seriously} don't want to underestimate
$t_0$!

One might reasonably choose to err by selecting a large $t_0$.
The rate of convergence is harder to estimate when $t_0$ is large, but
in general $\gamma_t$ itself forms a lower bound on the accuracy with which
the entropy may be known, with an unknown prefactor which is related to
the coherence time of the Monte Carlo simulation.  Since $\gamma_t$ is
given by $t_0/t$, the time to converge to a given accuracy is increased
in proportion to the ratio by which we overestimate $t_0$.  Thus, while
it is exponentially painful to underestimate $t_0$, overestimating by
several orders of magnitude is also not acceptable.  We should note
that these extreme limiting cases do not preclude the possibility that there
is a wide range of $t_0$ values that lead to an acceptable convergence
rate.

@@

\begin{figure}
  \includegraphics[width=\columnwidth]{gamma-n256}
  \caption{The update factor $\gamma_t$ versus iteration for $N=256$
    square-well atoms, for four different methods: WL, $1/t$-WL, SAMC,
    and SAD.}
    \label{fig:gamma-vs-t}
\end{figure}
Figure~\ref{fig:gamma-vs-t} compares $\gamma_t$ for the related
methods SAD, WL, $1/t$-WL, and SAMC.  For SAMC, $\gamma_t$ remains
constant before dropping as $1/t$.  WL $\gamma_t$ remains at 1 for
many iterations, and then decreases very rapidly, with $1/t$-WL
behaving similarly but decreasing more aggressively before
transitioning to a more conservative $1/t$ behavior.  The update
factor for SAD fluctuates dynamically around a value less than 1 for
early MC moves, and then decreases as approximately $1/t$ while
continuing to fluctuate as new energies are found to be important.  At
intermediate times, the SAD $\gamma_t$ decreases as $1/t^2$ before
asymptoting to $N_S/t$, which is the same as $1/t$-WL.

@@

A significant advantage of SAD over SAMC---which the $1/t$-WL and WL
methods share \emph{after} they have found all the energies---is that
the schedule for $\gamma_t$ automatically responds to the choice of bin
size. Although we have not yet tested this,
SAD should perform similarly over a wide range of bin sizes
because $\gamma_t \propto \nicefrac{N_S}{t}$.  As the number of energy
states $N_S$ found increases (fine binning), the time spent $t$ in
each bin will decrease with the effect that the convergence should be
roughly independent of the bin size chosen.  SAMC could be used with a
prefactor $\gamma_0$ to aid in a similar
way~\cite{werlich2015stochastic} but this adds yet another parameter
for the user to choose.

@@

\section{Conclusions}
We have introduced a new algorithm, a variant of the SAMC method, which
effectively samples the
energy space corresponding to a desired range of temperatures for a
variety of system sizes.
%
We find that both SAD and $1/t$-WL demonstrate excellent and robust
convergence.
They both converge more rapidly than SAMC, and unlike WL consistently
converge to the correct density of states. SAD requires
the user to specify a temperature range of interest rather than an
energy range of interest as $1/t$-WL does.  For use cases in which
a range of desired temperatures is known, this will make the SAD method
considerably more convenient.
